Supermartingales and Termination Analysis of Statistical PCF

### 2020-10-06 probabilisticProgrammingMartingales.tex
Continue proving canonicalCousins.
Write more cases, merge the forwards and backwards Y \sim_c X cases,
tweak the definition of \sim_c case f to make this proof work a bit
neater.

        \item $N \mid P = \tY (\lambda x. A)$, $A$ reduced at $Q$ is $A'$, $(R_i)_i$ is a list of all the positions where $A' \mid R  = x$, ordered from left to right, $O_1$ is $N$ reduced first at $P;\tY;\lambda;Q$ then at $P$, and $O_2$ is $N$ reduced first at $P$ then at $P;\lambda;@_1;Q$ then at $P;\lambda;@_1;R_i;Y;\lambda;Q$ for each $i$ in order.
    \end{enumerate}
    (in which case $O_1$ and $O_2$ are equal as terms, but with different reduction sequences), $O_1'$ and $O_2'$ are the results of applying some reduction sequence to each of $O_1$ and $O_2$ (the same reductions in each case, which is always possible because they're equal terms), and $S$ is a position in $O_1'$ (or equivalently $O_2'$), then $(O_1',S) \sim_c (O_2',S)$.
\end{enumerate}
The $\sim_c$-rules are illustrated in the following diagrams.
\end{definition}

@@

\item $Y \sim_c X$ in the forwards direction (so that $X$ is closer to CBV order than $Y$ is), and $\alpha$ comes after $\gamma$ in CBV order: In this case, the $\gamma$ and $\alpha$ reduction steps are already in the correct order in $Y$, therefore $\beta$ and $\gamma$ are already the last out-of-order pair of reductions in $Y$, $Y_1 = X$, and $Y \sim_c X \sim_c^* X_{cbv}$ is already the canonical sequence from $Y$ therefore $Y_{cbv} = X_{cbv}$.
\item $Y \sim_c X$ is forwards and $\alpha$ is disjoint from $\gamma$ and $\beta$: In this case, the sequence $(X_i)_i$ starts by swapping $\alpha$ and $\beta_1$ by case a, then because $\gamma$ is disjoint from $\alpha$ and comes after is in CBV order, $\gamma$ is right of $\alpha$, therefore either all the remaining $\beta_i$s, are $\geq \gamma$ (if $\beta$ and $\gamma$ aren't disjoint) or there are no remaining $\beta_i$s (if $\beta$ and $\gamma$ are disjoint, in which case they swap by case a and there is only $\beta_1$). The sequence $(X_i)_i$ therefore proceeds to swap all the remaining $\beta_i$s then $\gamma$, in order, below $\alpha$, resulting in $X_i$ for some $i$ having as its relevant sub-reduction-sequence $\alpha, \gamma, \beta_k, \dots, \beta_1$. The other sequence $(Y_j)_j$ starts by swapping $\gamma$ then $beta$ below $\alpha$, then swapping $\gamma$ and $\beta$. Because $\alpha$ is disjoint from $\gamma$, the reduction there does not affect the sub-skeleton at position $\gamma$, therefore this last $\sim_c$ step proceeds identically to how it did in $Y \sim_c X$, resulting in $\gamma, \beta_k, \dots, \beta_i$. Overall, this results in $Y_3$ having the relevant sub-reduction-sequence $\alpha, \gamma, \beta_k, \dots, \beta_1$, therefore $Y_3 = X_i$ for the aforementioned value of $i$.
\item $Y \sim_c X$ is backwards and $\alpha$ is disjoint from $\beta$ and $\gamma$: In this case, $(X_i)_i$ proceeds by swapping $\beta$ then $\gamma$ below $\alpha$ by case a, then reversing $\gamma$ and $\beta$ back to $\beta, \gamma_k, \dots, \gamma_i$ by the same case as in $Y \sim_c X$ but in the other direction, and $(Y_j)_j$ proceeds by swapping each $\gamma_i$ in turn and then $\beta$ below $\alpha$ by case a, so both sequences at some point reach the same result, with $\alpha, \beta, \gamma_k, \dots, \gamma_1$ as the relevant sub-reduction-sequence.

@@

\item $Y \sim_c X$ backwards, $\alpha \leq \beta$ and both $\alpha$ and $\beta$ are left of $\gamma$: This case is quite similar to the previous one. In this case, $(X_i)_i$ proceeds by swapping $\beta$ below $\alpha$ by one of the cases b to f, then swapping $\gamma$ below $\alpha$ then each of the images of $\beta$ in turn by case a, and $(Y_j)_j$ proceeds by swapping $\gamma$ below $\alpha$ by case a, then swapping $\beta$ below $\alpha$. As before, the results of the swapping of $\alpha$ and $\beta$ are the same in both cases, therefore both sequences reach the same reduction sequence containing $\alpha$ then the images of $\beta$ then $\gamma$.

@@

  \item $\gamma, \beta_k, \dots, \beta_1$ has length 2, $\alpha < \beta_1$ and $\gamma$ is right of $\alpha$: In this case, the sequence $(X_i)_i$ starts by swapping $\alpha$ and $\beta_1$ by one of the cases b to f in the definition of $\sim_c$, then swapping $\gamma$ down past $\alpha$ by case a, then down past all the images of $\beta_1$, also by case a. Eventually it reaches some $X_i$ where the relevant sub-reduction-sequence is $\alpha$ then whatever $\beta_1$ was turned into in $X \sim_c X_1$ then $\gamma$, before moving on to swapping pairs earlier in the reduction sequence. The sequence $(Y_j)_j$ starts by swapping $\gamma$ down past $\alpha$ by case a, then swapping $\alpha$ and $\beta_1$ as above. The subskeleton at $\alpha$ at the relevant point in the reduction sequence is the same as in the $(X_i)_i$ case because $\gamma$ and $\alpha$ are disjoint therefore it's unaffected by the reduction at $\gamma$, therefore the swap of $\alpha$ and $\beta_1$ results in the same sub-reduction-sequence as in the $(X_i)_i$ case, therefore $Y_2 = X_i$ (with $i$ taking the value mentioned earlier where the rearrangement of this sub-reduction-sequence is done), therefore $X_{cbv} = Y_{cbv}$.
    \item \akr{The cases below this point aren't wrong, they're just not such a neat way to split up the possible cases so I probably won't be using them.}
    \item $\gamma, \beta_k, \dots, \beta_1$ has length 2 and $\alpha < \beta_1$ as before, but now $\gamma$ is left of $\alpha$: In this case, $(Y_j)_j$ starts by swapping $\beta_1$ and $\gamma$, therefore $Y_1 = X$ already.
    \item $\gamma, \beta_k, \dots, \beta_1$ has length 2 and $\alpha < \beta_1$ as before, but now $\gamma < \alpha$: Again, $(Y_j)_j$ starts by swapping $\beta_1$ and $\gamma$ therefore $Y_1 = X$.
    \item $\gamma, \beta_k, \dots, \beta_1$ has length 1 (so that $\gamma = \beta_1$): In this case, $Y \sim_c X$ either by case b or by case e (where the lambda has no occurrences of its variable), therefore $\gamma < \delta$ (the position of the reduction which is essentially deleted in $Y \sim_c X$). $\alpha$ is either outside or left of $\gamma$, therefore it bears the same relation to $\delta$. The sequence $(X_i)_i$ starts by swapping $\gamma$ below $\alpha$, and $(Y_j)_j$ starts by swapping $\gamma$ then $\delta$ below $\alpha$, then what happens next depends on the shape of the term at $\alpha$, or equivalently, by which case of $\sim_c$ are $\gamma$ and $\delta$ swapped below $\alpha$. If it's cases a, c or d, each of $\gamma$ and $\delta$ corresponds to one reduction in $Y_2$, and the relative position of these is the same as the relative position of $\gamma$ and $\delta$ therefore the next step of $(X_i)_i$ involves these two reductions and deletes the one equivalent to $\delta$, so $Y_3 = X_1$. If it's case b or e, in $X_2$ there are $n$ reductions corresponding to $\delta$ and $n$ reductions corresponding to $\gamma$, and they are all disjoint except that for each image of $\delta$, there's one image of $\gamma$ that it's inside, in a position that is deleted by the reduction at that image of $\gamma$. The next $\frac{n(n+1)} 2$ steps in $(X_i)_i$ are then each image of $\delta$ in turn being swapped (by case a) past the images of $\gamma$ until they reach the one they match, at which point they get deleted. In this case therefore, $Y_{n(n+1)/2+2} = X_1$. Case f, corresponding to $\tY$-reduction, is more complicated because the images of $\gamma$ and $\delta$ are not all disjoint. For each of $\gamma$ and $\delta$, there is one image of it that may contain other such positions. Let these be $\gamma_o$ and $\delta_o$. The sequence $(Y_j)_j$ then proceeds by moving all of the images of $\delta$ which are right of $\gamma_o$ to their corresponding $\gamma$-image and deleting them, then moving all the images of $\delta$ which are after $\delta_o$ but which are inside $\gamma_o$ (one at a time) past all the images of $\gamma$ left of $\gamma_o$ by case a, then either deleting them against $\gamma_o$ by case b or e as appropriate, or moving them past it by case c or d, then possibly past some of the images of $\gamma$ inside $\gamma_o$, then deleting them once they reach the matching image of $\gamma$. Then $\delta_o$ itself is moved to $\gamma_o$ then deleted, then the remaining images of $\delta$ are similarly moved and deleted. The end result of this process is that all the images of $\delta$ are deleted and all the images of $\gamma$ are unaffected, therefore for some $j$, $Y_j = X_1$.
    \item $\alpha$ is disjoint from $\gamma$ and all the $\beta_i$: In this case, $\alpha$ is left of all the $\beta_i$, and if it isn't also left of $\gamma$, then $Y_1 = X$ and we're done. Otherwise, $(X_i)_i$ proceeds by swapping each of the $\beta_i$ in turn then $\gamma$ below $\alpha$ by case a, and $(Y_j)_j$ swaps everything up to $\delta$ to below $\alpha$ by case a, then performs the same swap as in $Y \sim_c X$, but after $\alpha$ in the reduction sequence, yielding $\alpha, \gamma, \beta_k, \dots, \beta_1$, which matches $X_{k+1}$.
    \item $\alpha > \gamma$: In this case, $Y_1 = X$.
    \item \akr{There are several more cases to go, and that last case is more typical of how complex they'll be. I'd guess around 5 more, but I haven't worked them all out precisely. Also, I might find a neater way to deal with case f, or at least not have to repeat that argument, because the case where both reductions are $\tY$-reductions should include all the complexity of any other case involving case f.}
\end{itemize}

\paragraph{}
If $Y$ is in CBV order and $Y \sim_c^* X$, then $Y_{cbv} = X_{cbv}$ but also $Y_{cbv} = Y$ therefore $X_{cbv}$ is unique.
\end{proof}

>>>>

        \item $N \mid P = \tY (\lambda x. A)$, $A$ reduced at $Q$ is $A'$, $(R_i)_i$ is a list of all the positions where $A' \mid R  = x$, ordered from left to right, $O_1$ is $N$ reduced first at $P;\tY;\lambda;Q$ then at $P$, and $O_2$ is $N$ reduced first at $P$ then at $P;\lambda;@_1;R_i;Y;\lambda;Q$ for each $i$ in order where $R_i$ is left of $Q$ then at $P;\lambda;@_1;Q$ then at $P;\lambda;@_1;R_i;Y;\lambda;Q$ for the remaining values of $i$.
    \end{enumerate}
    (in which case $O_1$ and $O_2$ are equal as terms, but with different reduction sequences), $O_1'$ and $O_2'$ are the results of applying some reduction sequence to each of $O_1$ and $O_2$ (the same reductions in each case, which is always possible because they're equal terms), and $S$ is a position in $O_1'$ (or equivalently $O_2'$), then $(O_1',S) \sim_c (O_2',S)$.
\end{enumerate}
The $\sim_c$-rules are illustrated in the following diagrams.
\end{definition}

@@

  \item $Y \sim_c X$ in the backwards direction (so that $Y$ is closer to CBV order than $X$ is): By swapping the roles of $X$ and $Y$, and $\beta$ and $\gamma$, this is equivalent to the $Y \sim_c X$ forwards case. The assumption that $\alpha$ is before $\beta_1$ in CBV order (so that $\beta_1, \alpha$ is the pair to be swapped in $X \sim_c X_1$) maps to the assumption that $\alpha$ is before $\gamma_1$ (so that the result doesn't just follow trivially by $Y_1$ being equal to $X$) and vice-versa. Then one of the cases below establishes that $X_i = Y_j$ for some $i$ and $j$, therefore after swapping back, $Y_i = X_j$.
\item $Y \sim_c X$ forwards and $\alpha$ comes after $\gamma$ in CBV order: In this case, the $\gamma$ and $\alpha$ reduction steps are already in the correct order in $Y$, therefore $\beta$ and $\gamma$ are already the last out-of-order pair of reductions in $Y$, $Y_1 = X$, and $Y \sim_c X \sim_c^* X_{cbv}$ is already the canonical sequence from $Y$ therefore $Y_{cbv} = X_{cbv}$.
\item $Y \sim_c X$ forwards and $\alpha$ is disjoint from $\gamma$ and $\beta$: In this case, the sequence $(X_i)_i$ starts by swapping $\alpha$ and $\beta_1$ by case a, then because $\gamma$ is disjoint from $\alpha$ and comes after is in CBV order, $\gamma$ is right of $\alpha$, therefore either all the remaining $\beta_i$s, are $\geq \gamma$ (if $\beta$ and $\gamma$ aren't disjoint) or there are no remaining $\beta_i$s (if $\beta$ and $\gamma$ are disjoint, in which case they swap by case a and there is only $\beta_1$). The sequence $(X_i)_i$ therefore proceeds to swap all the remaining $\beta_i$s then $\gamma$, in order, below $\alpha$, resulting in $X_i$ for some $i$ having as its relevant sub-reduction-sequence $\alpha, \gamma, \beta_k, \dots, \beta_1$. The other sequence $(Y_j)_j$ starts by swapping $\gamma$ then $beta$ below $\alpha$, then swapping $\gamma$ and $\beta$. Because $\alpha$ is disjoint from $\gamma$, the reduction there does not affect the sub-skeleton at position $\gamma$, therefore this last $\sim_c$ step proceeds identically to how it did in $Y \sim_c X$, resulting in $\gamma, \beta_k, \dots, \beta_i$. Overall, this results in $Y_3$ having the relevant sub-reduction-sequence $\alpha, \gamma, \beta_k, \dots, \beta_1$, therefore $Y_3 = X_i$ for the aforementioned value of $i$.

@@

@@

\item $Y \sim_c X$ forwards, $\alpha;@_1 < \gamma$, and $\alpha;@_2 < \beta$: Let $x$ be the variable involved in the $\beta$-reduction at $\alpha$, so that the sub-skeleton at $\alpha$ is $(\lambda x. A) B$ for some $A, B$, where $\gamma = \alpha;@_1;\lambda;\gamma'$ and $A \to A'$ at $\gamma'$, and $\beta = \alpha;@_2;\beta'$, and $B \to B'$ at $\beta'$. In this case, $(X_i)_i$ starts by swapping $\beta$ below $\alpha$ by case e, producing one image of $\beta$ for each $x$ in $A'$, then swapping $\gamma$ below $\alpha$ by case d, producing a single image $\alpha;\gamma'$. For each of the instances of $x$ in $A$ left of $\gamma'$, the reduction at $\alpha;\gamma'$ is then swapped below the corresponding image of $\beta$. The sequence $(Y_j)_j$ starts by swapping $\gamma$ below $\alpha$ then $\beta$ below $\alpha$, but in this case the images of $\beta$ produced may be different. In this case the swap happens earlier in the reduction sequence than $\gamma$, so that the body of the lambda is still $A$ rather than $A'$, and the reduction at $\gamma$ may rearrange or change the number of instances of $x$. Next, each of the images of $\beta$ to the right of $\alpha;\gamma'$ are swapped with it by case a. At this point, the next few reductions before $\alpha;\gamma'$ will in general be images of $\beta$ at positions inside $\alpha;\gamma'$. The subskeleton at position $\alpha;\gamma'$ before these reductions is $(A|\gamma')[B/x]$, then by the images of $\beta$ this reduces to $(A|\gamma')[B'/x]$, then it reduces at its root position ($\alpha;gamma'$ in the overall skeleton) to $(A'|\gamma')[B'/x]$. For each position of an $x$ in $(A|\gamma')$, there are 0 or more corresponding positions of $x$ in $(A'|\gamma')$, depending on what type of reduction $A \to A'$ is. Because $B$ and $B'$ cannot contain any instances of the variable involved in the reduction $A \to A'$ (if it's a $\beta$-reduction or $\tY$-reduction), each instance of $B$ in $(A|\gamma')[B/x]$ has the same set of images in $(A'|\gamma')[B/x]$ as the corresponding $x$ in $(A|\gamma')$. When these images of $\beta$ that overlap with the image of $\gamma$ are swapped below it, each of them has its own images, one at each position of $B$ in $(A'|\gamma')[B/x]$ corresponding to its original copy of $B$ in $(A|\gamma')[B/x]$. After all of them are swapped below $\alpha;\gamma'$ (and rearranged among themselved by case a), there is therefore one reduction at each copy of $B$ in $(A'|\gamma')[B/x]$, i.e.~one at position $\beta'$ relative to each $x$ in $(A'|\gamma')$. They are therefore the same as the images of $\beta$ in $X_1$ that overlap with $\alpha;\gamma'$, because those are also one reduction at position $\beta'$ relative to each $x$ in $(A'|\gamma)$. Both $(X_i)_i$ and $(Y_j)_j$ therefore eventually reach a point where the relevant sub-reduction-sequence is $\alpha$, the images of $\beta$ to the left of $\alpha;\gamma'$, $\alpha;\gamma'$ itself, the aforementioned images of $\beta$ that overlap with $\alpha;\gamma'$, then all the images of $\beta$ to the right of $\alpha;\gamma'$.
\item $Y \sim_c X$ forwards, $\gamma$ and $\beta$ are disjoint and both $> \alpha$, and the reduction at $\alpha$ is a $\tY$-reduction: This is similar to the previous case, but more complicated, so it will need some definitions to explain properly what's going on. Let the term at $\alpha$ before any of the reductions be $\tY \lambda x. A$, let $\gamma = \alpha;\tY;\lambda;\gamma'$, let $\beta = \alpha;\tY;\lambda;\beta'$, let $A | \beta' = B \to B'$ with the reduction at $\cdot$, let $A | \gamma' = C \to C'$ with the reduction at $\cdot$, let the positions in $A$ where $x$ occurs that are left of $\gamma'$, between $\gamma'$ and $\beta'$ (but still disjoint from both), and right of $\beta'$ respectively be $(\delta^l_i)$, $(\delta^m_i)$ and $(\delta^r_i)$, let the positions in $B$, $C$, $B'$ and $C'$ respectively where $x$ occurs be $(\delta^B_i)$, $(\delta^C_i)$, $(\delta^{B'}_i)$ and $(\delta^{C'}_i)$ (so that all the positions in $A$ where $x$ occurs, in left-to-right order, are $(\delta^l_i)_i, (\delta^C_i)_i, (\delta^m_i)_i, (\delta^B_i)_i, (\delta^r_i)_i$). The sequence $(X_i)_i$ starts by swapping $\beta$ then $\gamma$ below $\alpha$. With the shorthand that $\gamma'(\epsilon) = \alpha;\lambda;@_1;\epsilon;\tY;\lambda;\gamma'$, $\gamma'_0 = \alpha;\lambda;@_1;\gamma'$, and similarly for $\beta'$, the relevant portion of $X_2$ is then $\alpha,\ (\gamma'(\delta^l_i))_i,\ \gamma'_0,\ (\gamma'(\delta^{C'}_i))_i,\ (\gamma'(\delta^m_i))_i,\ (\gamma'(\delta^B_i))_i,\ (\gamma'(\delta^r_i))_i,\ (\beta'(\delta^l_i))_i,\ (\beta'(\delta^{C'}_i))_i,\ (\beta'(\delta^m_i))_i,\ \beta'_0,\ (\beta'(\delta^{B'}_i))_i,\ (\beta'(\delta^r_i))_i$. Next in $(X_k)_k$, for each $i$, $\gamma'(\delta^r_i)$ are swapped past all the images of $\beta$ until it is immediately before $\beta'(\delta^r_i)$ by case a, then for each $i$, $\gamma'(\delta^B_i)$ is swapped past all the images of $\beta$ until $\beta'_0$ by case a, then swapped with $\beta'_0$ by one of the other cases depending on what type of reduction $B \to B'$ is and the relative positions. Even if it's case d or f, all of the images of $\gamma'(\delta^B_i)$ are disjoint from each other (and from the positions of the other reductions after $\beta'_0$) because the redex, $C$, doesn't contain any instances of the variable involved in the reduction $B \to B'$. Expanding the definitions, this is swapping $\alpha;\lambda;@_1;\delta^B_i;\tY;\lambda;gamma'$ with $\alpha;\lambda;@_1;\beta'$ at a point in the reduction sequence where the subskeleton at $\alpha;\lambda;@_1;\beta'$ is $B$ with a mixture of $\tY \lambda x. A$ and $\tY \lambda x. A[C'/\gamma']$ substituted for its occurrences of $x$, and one of these occurrences of $x$ is at $\delta^B_i$. For each $\delta^B_j$, there are some corresponding $\delta^{B'}_k$, where the images of the $x$ at $\delta^B_j$ end up after the reduction $B \to B'$. These cover all the $\delta^{B'}_k$, and uniquely, and for each $\delta^B_i$, the images of $\zeta;\delta^B_i;\theta$ after swapping it with $\zeta;\beta'$ are precisely $(\zeta;\delta^{B'}_k;\theta)$ for those same values of $k$, therefore after swapping all of the $(\gamma'(\delta^B_i))_i$ down past $\beta'_0$, their images are $(\gamma'(\delta^{B'}_i))_i$. After swapping with $\beta'_0$, each of these images than swaps by case a to take its place among the $(\beta'(\delta^{B'}_i))_i$. Next up in $(X_i)_i$, the $(\gamma'(\delta^m_i))_i$s and the $(\gamma'(\delta^{C'}_i))_i$s swap down past some images of $\beta$ they're disjoint from to take their place before their matching image of $\beta$, then $\gamma'_0$ swaps past the $(\beta'(\delta^l_i))_i$, stopping immediately before $\beta'(\delta^{C'}_1)$, then the $(\gamma'(\delta^l_i))_i$ and the $(\beta'(\delta^l_i))_i$ mix. The end result of the rearrangement of (this subsequence of) the reduction sequence is therefore $\alpha,\ (\gamma'(\delta^l_i), \beta'(\delta^l_i))_i,\ \gamma'_0,\ (\gamma'(\delta^{C'}_i), \beta'(\delta^{C'}_i))_i,\ (\gamma'(\delta^m_i), \beta'(\delta^m_i))_i,\ \beta'_0,\ (\gamma'(\delta^{B'}_i), \beta'(\delta^{B'}_i))_i,\ (\gamma'(\delta^r_i), \beta'(\delta^r_i))_i$.

  The other sequence, $(Y_j)_j$, proceeds similarly, but with the images of $\gamma$ starting out after the images of $\beta$. It swaps all the images of $\beta$ that are right of $\gamma_0$ to the appropriate places by case a, then swaps all the $(\beta'(\delta^C_i)_i$ first to then past $\gamma'_0$, resulting in $(\beta'(\delta^{C'}_i))_i$ for the same reason as with swapping those images of $\gamma$ that overlapped with $\beta'_0$ past it in the $(X_i)_i$ case above, then finally swapping the $(\beta'(\delta^l_i))_i$s and the $(\gamma'(\delta^l_i))_i$s into the correct order. As required, this produces the same result as in $(X_i)_i$.
  \item $Y \to X$ forwards, and $\alpha < \gamma < \beta$: \akr{This is the final case, although I may have to split it up depending on whether the reduction at $\alpha$ is a $tY$-reduction.}
\end{itemize}

\paragraph{}
If $Y$ is in CBV order and $Y \sim_c^* X$, then $Y_{cbv} = X_{cbv}$ but also $Y_{cbv} = Y$ therefore $X_{cbv}$ is the unique reduction sequence in call-by-value order that's related to $X$ by $\sim_c^*$.
\end{proof}
